### 2016-06-11 abstract.tex
intro and abstract

\begin{abstract}
ACID model of database transactions advocates serializability as the
correctness criterion for concurrent transactions. 
% Serializability is an easily-stated property,
% providing database programmers with a simple
% implementation-independent model to reason about program invariants in
% presence of concurrent transactions. Unfortunately, 
While serializability is an easily-stated high-level property,
enforcing it requires the database to severely restrict
concurrency.
% , thus limiting the availability and performance
% (latency/throughput) of the data store.  In the database parlance,
% serializibility is \emph{not highly available}~\cite{bailishotos13},
% making it particularly unsuitable for modern-day distributed web
% applications that aspire to be \emph{always on}.
Pragmatic concerns led to the development of various weaker
alternatives to serializability, which are overwhelmingly preferred in
practice.
% some of which have been
% standardized by the community as \emph{ANSI SQL isolation
% levels}~\cite{berenson}. Indeed, weakly isolated transactions are
% overwhelmingly preferred to serializability in practice, to the extent
% that the default isolation level in popular industrial-strength
% databases is often weaker than serializability.
However, unlike serializability, the semantics of weaker isolation
levels are often described informally in terms of an implementation.
While there have been some attempts at formalization, there
nonetheless exists a wide chasm between formal models of weak
isolation and the operational models of programs, hindering uniform
reasoning about program invariants in presence of weakly isolated
transactions.

In this paper, we propose an operational model and a compositional
proof system for concurrent imperative programs with weakly isolated
transactions (\txnimp programs, as we call them).  Our operational
model captures all the behaviors exhibited by a program under weak
isolation without referring to any specific implementation of
isolation. Our proof system enables formal reasoning about program
invariants while taking into account the fine-grained isolation
semantics of transactions. Case studies demonstrate the utility of our
approach.
\end{abstract}

>>>>

\begin{abstract}

  Serializability is a well-understood mechanism that simplifies
  reasoning about the behavior of concurrent transactions.  However,
  its two key properties, \emph{atomicity} - a transaction's effects
  are seen only when it commits, and \emph{isolation} - a transaction
  never witnesses the effects of other committed transactions while it
  is executing, impose significant constraints on implementations, and
  can negatively impact database performance by restricting
  opportunities to exploit concurrency.

  As a result, database systems (in practice) support, and often
  encourage, developers to use weaker alternatives.  Generally, these
  alternatives break the strong isolation guarantees provided by
  serializable transactions to permit greater concurrency.
  Unfortunately, the semantics of weak isolation is often poorly
  understood, and usually explained informally in terms of low-level
  implementation artifacts.  Consequently, verifying high-level
  correctness properties in such environments can be problematic.

  To address this issue, this paper presents a novel program logic
  that enables compositional reasoning about the behavior of
  concurrently executing weakly-isolated transactions.  Notably, our
  development is parametric over a transaction's specific isolation
  semantics, and the consistency guarantees provided by the underlying
  data store, allowing it to be applicable over a large range of
  different isolation and consistency models.  Case studies and
  experiments on real-world applications demonstrate the utility of
  our approach, and provide strong evidence that weakly-isolated
  transactions can be placed on the same formal footing as their
  strongly-isolated serializable counterparts.

\end{abstract}

### 2016-06-11 introduction.tex
intro and abstract

Database transactions allow users to group operations on multiple
objects into a coherent unit with useful properties including
atomicity, consistency, (serializable) isolation and durability
(ACID). ACID has been the gold standard for database transactions for
many decades, providing programmers with a simple
implementation-independent model to reason about concurrent
transactions. However, serializability, the ACID-prescribed standard for
isolation, does not come for free; it requires databases to rely on
expensive techniques such as two-phase locking~\cite{twopl,ullmanbook}
that severely restrict concurrency while incurring an additional
overhead of deadlock detection, rollback and re-execution. Enforcing
serializability in a replicated data store requires global
coordination among geo-distributed replicas, which makes the store
\emph{unavailable} in presence of the inevitable network
partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}. The
conflict between serializability and pragmatic goals of high
peformance and availability has motivated the development of multiple
weaker forms of transaction isolation beginning as early as
1976~\cite{gray1976}. ANSI SQL 92 standard defines three such weak
isolation \emph{levels} which are now implemented in
many relational and NoSQL databases. Weakly isolated transactions have
been found to significantly outperform serializable transactions on
benchmark suites, both on single-node databases and multi-node
replicated stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to
their overwhelming adoption. A 2013 study~\cite{bailishotos} of 18
popular ACID and ``NewSQL'' databases found that only 3 of them offer
serializability by default, and half, including Oracle 11g, do not
offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
database applications finds no evidence of any of them changing the
default isolation level offered by the database. Together, the studies
establish the prevalance of weak isolation in practice.

Unfortunately, weak isolation introduces new set of behaviors into
programs that often give rise to anomalous executions and
incomprehensible results~\cite{pldi15}. To quantify weak isolation
anomalies, Fekete et al~\cite{feketevldb09} have devised and
experimented with a microbenchmark suite that executes transactions
under \iso{Read Committed} isolation level - the default level for 8
of the 18 databases studied in~\cite{bailishotos}, and found that 25
out of every 1000 rows in the database violate at least one integrity
constraint. Bailis et al~\cite{bailisferal} rely on Rails's
\emph{uniqueness validation} to maintain uniqueness of records while
serving Linkbench's~\cite{linkbench} insertion workload (6400 records
distributed over 1000 keys; 64 concurrent clients), and end up with
more than 10 duplicate records. Rails relies on database transactions
to validate uniqueness during insertions, which is ok if transactions
are serializable, but incorrect under \iso{Read Committed} isolation
used in the experiments. The same study has found that 13\% of all
invariants among 67 open source Ruby-on-Rails applications are the
risk of being violated due to weak isolation. Indeed, incidents of
safety violations due to weak isolation have been reported on web
services in production~\cite{starbucksbug, scimedbug}, including in
safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. Evidently, the ill effects
of weak isolation are as prevalent as the weak isolation itself.

The root cause of problems with weak isolation is that its semantics
in context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details, such as the nature and
duration of locks held in each case. ANSI SQL 92 standard four levels
of isolation (including serializability) in terms of sevaral
undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is required
to prevent. While this is an improvement, it requires programmers to
be clairvoyant about the possible ways various undesirable phenomena
might manifest in their applications, and determine, in each case, if
the phenomenon can be allowed without violating application
invaraints. Adya's thesis~\cite{adyaphd} includes first formal
definitions of some of the well-known isolation levels in the context
of a sequentially consistent (SC) database. However, there have been
no attempts to relate Adya's system model to an operational model of
programs in a reasoning framework that facilitates formal proofs of
correctness. For this reason, the unsystematic way of reasoning about
weak isolation in terms of undesirable phenomena remains popular to
date, with major database vendors continuing to document their
isolation levels in terms of the such phenomena~\cite{postgresiso,
mysqliso, oracleiso}.

There have been encouraging developments in Programming Languages and
Systems research towards reasoning about application invariants in
presence of weak consistency~\cite{burckhardt14, redblueosdi,
redblueatc, ecinec, gotsmanpopl16}. Weak isolation can make use of
some of the reasoning inventory built for weak consistency (as we
demonstrate in later sections), but the former problem is strictly
more general than the latter. At a high level, reasoning about weak
consistency involves reasoning about concurrent execution of atomic
and fully isolated operations against different but consistent states
of the database, where each operation locally preserves invariants. In
contrast, reasoning about weak isolation involves reasoning about
concurrent execution of transactions, whose operations witness
different states which, although consistent individually, may not be
reconciled into a consistent view. Moreover, operations may violate
invariants as long as the transaction as a whole preserves them. It is
easy to see that the latter problem reduces to the former if all
transactions are singletons. The generality of weak isolation is also
evident from the fact that weakly isolated transactions have existed
on sequentially consistent relational databases long before the advent
of weakly consistent replicated stores. Thus, there is a need for
reasoning frameworks for weak isolation that generalize those for weak
consistency in some important ways. The framework should be general
enough to reason about the semantics of various isolation levels,
including ANSI SQL isolation levels, in the context of data stores
with various consistency semantics (\eg sequentially consistent data
store of~\cite{adyaphd}, causally consistent store assumed
in~\cite{gotsmanpopl16} etc). Considering the increasing diversity in
the semantics of weak consistency~\cite{zoo} and
isolation~\cite{gotsmanconcur}, it is ideal if the reasoning framework
is parameterized on the semantics of consistency and isolation.  In
short, what is neededis a reasoning framework that can answer the
following question:
\begin{center}
\emph{Given a program annotated with desired invaraints, and isolation
semantics for transactions in the program, assuming that the program
executes against a data store with a given consistency semantics, are
the invariants guaranteed to hold?}
\end{center}

In this paper, we propose a reasoning framework that attempts to
answer this question.

>>>>

Database transactions allow users to group operations on multiple
objects into a coherent unit, equipped with a set of four key
properties - atomicity, consistency, isolation, and durability
(ACID). ACID has been the gold standard for database transactions for
many decades, providing programmers with a simple implementation
independent model to reason about concurrent transactions.  For
example, concurrency control mechanisms typically enforce isolation by
making all transactions \emph{serializable}, to ensure that any
allowed schedule is equivalent to one in which there is no
interleaving of actions from different concurrently executing
transactions.

However the strong guarantees provided by serializability does not
come for free - pessimistic concurrency control methods, for example,
require databases to provide expensive techniques such as two-phase
locking that incur overhead to deal with deadlocks, rollbacks, and
re-execution~\cite{twopl,ullmanbook}.  Similar criticisms apply to
optimistic multi-version concurrency control methods that must deal
with timestamp and version management~\cite{BG81}.  Regardless of the
method employed, enforcing serializability in a replicated data store
additionally requires global coordination among geo-distributed
replicas, which makes the store \emph{unavailable} in the presence of
inevitable network
partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.

The tension between serializable transactions, which are easy to
reason about but difficult to implement, and more pragmatic goals
driven by performance and availability considerations, has motivated
the development of multiple weaker forms of transaction isolation,
beginning as early as 1976~\cite{gray1976}. The ANSI SQL 92 standard
defines three such weak isolation \emph{levels} which are now
implemented in many relational and NoSQL databases. Not surprisingly,
weakly isolated transactions have been found to significantly
outperform serializable transactions on benchmark suites, both on
single-node databases and multi-node replicated
stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to their
overwhelming adoption. A 2013 study~\cite{bailishotos} of 18 popular
ACID and ``NewSQL'' databases found that only three of them offer
serializability by default, and half, including Oracle 11g, do not
offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
database applications finds no evidence that applications manifestly
change the default isolation level offered by the database. Taken
together, these studies make clear that weakly isolated transactions
are quite prevalent in practice.

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} have devised and experimented
with a microbenchmark suite that executes transactions under a weakly
isolated \iso{read committed} isolation level - the default level for
8 of the 18 databases studied in~\cite{bailishotos}, and found that 25
out of every 1000 rows in the database violate at least one integrity
constraint. Bailis \empH{et al.}~\cite{bailisferal} rely on Rails's
\emph{uniqueness validation} to maintain uniqueness of records while
serving Linkbench's~\cite{linkbench} insertion workload (6400 records
distributed over 1000 keys; 64 concurrent clients), and end up with
more than 10 duplicate records. Rails relies on database transactions
to validate uniqueness during insertions, which is ok if transactions
are serializable, but incorrect under \iso{Read Committed} isolation
used in the experiments. The same study has found that 13\% of all
invariants among 67 open source Ruby-on-Rails applications are at risk
of being violated due to weak isolation. Indeed, incidents of safety
violations due to weak isolation have been reported on web services in
production~\cite{starbucksbug, scimedbug}, including in
safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.

A major problem with weak isolation is that its semantics in the
context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details, such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard four
levels of isolation (including serializability) in terms of sevaral
undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is required
to prevent. While this is an improvement, it requires programmers to
be prescient about the possible ways various undesirable phenomena
might manifest in their applications, and determine, in each case, if
the phenomenon can be allowed without violating application
invariants. This is understandably hard. Adya~\cite{adyaphd} includes
the first formal definitions of some of well-known isolation levels in
the context of a sequentially consistent (SC) database.  However,
there has been little progress relating Adya's system model to a
formal operational semantics or a proof system that can facilitate
rigorous correctness arguments.  Consequently, reasoning about weak
isolation remains an error prone endeavor, with major database vendors
continuing to document their isolation levels primarily in terms of
the undesirable phenomena~\cite{postgresiso, mysqliso, oracleiso} a
particular isolation level may induce; it thus squarely becomes the
programmer's responsibility to determine if their application can
yield an execution which can trigger such phenomena.

Recent results on reasoning about application invariants in the
presence of weak consistency~\cite{burckhardt14, redblueosdi,
  redblueatc, ecinec, gotsmanpopl16} provide insights on possible ways
we might overcome some of the difficulties in understanding weak
isolation, even though the latter is strictly more general than the
former.  At a high level, reasoning about weak consistency involves
reasoning about the concurrent execution of atomic and strongly
isolated operations against different but consistent states of the
database, where each operation locally preserves application
invariants. In contrast, reasoning about weak isolation involves
reasoning about concurrent execution of transactions, whose operations
witness different states, which, although consistent individually, may
not be obviously reconciled into a consistent view. Moreover,
operations may violate invariants as long as the transaction as a
whole preserves them.  Intuitively, the latter problem reduces to the
former if all transactions contain a single operation. Furthermore,
weak isolation can exist independent of weak consistency, as evident
from the presence of weakly isolated transactions on sequentially
consistent relational databases; such systems existed long before the
advent of weakly consistent NoSQL stores.  Thus, frameworks for
reasoning about weak isolation will necessarily have to generalize the
reasoning frameworks developed for weak consistency in important ways.

% The framework should be general enough to reason about the semantics
% of multiple isolation levels, including those proposed after the SQL
% 92 standard, in the context of various stores (\eg sequentially
% consistent store of~\cite{adyaphd}, causally consistent store
% of~\cite{gotsmanpopl16} etc).

In this paper, we propose a program logic for weakly isolated
transactions that realizes this goal.  In particular, we develop a set
of syntax-directed compositional proof rules that allow developers to
build correctness proofs for transactional programs in the presence of
a weakly isolated concurrency control mechanism.  A key novelty of our
approach is that it is parametric over the \emph{isolation semantics}
of transactions in the program, as well as the \emph{consistency
  semantics} of the store. In concrete terms, this means that, unlike
recent work focussed on formalizing the semantics of weak
consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system model
does not assume a minimum consistency/isolation level, nor does it
assume a predefined set of such levels.  Instead, our operational
semantics admits declarative specifications of transaction isolation
and store consistency levels (\eg~\cite{pldi15,gotsmanconcur15}), and
generates executions that are guaranteed to satisfy these
specifications. The result is a flexible system model that is general
enough to incorporate the semantics of a range of isolation levels on
a variety of stores (\eg the sequentially consistent store
of~\cite{adyaphd}, or causally consistent store
of~\cite{gotsmanpopl16}, etc).  Our key technical contribution is thus
a set of proof rules that demonstrate that a program with a given
selection of isolation levels for its transactions preserves its
invariants when executed on a store satisfying certain consistency
guarantees.

The paper makes the following contributions:
\begin{itemize}
  \item We develop a semantics for a core language equipped with
    weakly isolated transactions.  Building on some ideas espoused in
    previous work, we show that isolation semantics can be captured as
    well-formedness constraints on a program's execution.
  \item We propose a compositional proof system for this language
    capable of relating high-level application invariants to the structure
    of traces induced by the operational semantics, in which transactions
    are parameterized with specific weak isolation levels.
  \item We define a \emph{maximum visibility principle} to reconcile
    store consistency guarantees with transaction isolation
    requirements, and use it to instantiate our operational model and
    the proof system for various kinds of stores.
  \item Case studies modeled after real-world scenarios demonstrate
    the applicability and utility of our proof methodology.
\end{itemize}

Our semantic framework and proof methodology thus provide the first
(to the best of our knowledge) mechanism that precisely relates
high-level program invariants to low-level store consistency and weak
isolation guarantees, thereby allowing weakly isolated transactions to
enjoy the same rigorous reasoning capabilities as their strongly
isolated (serializable) counterparts.
